---
title: "Reproducible Research: Peer Assessment 1"
author: "Vladimir Sazontov"
output: html_document
---

## Loading and preprocessing the data

The original data is loaded as follows:
```{r}
url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, destfile = "repdata-data-activity.zip", mode="wb")
raw_data <- read.csv(unzip("repdata-data-activity.zip"), header = TRUE)
```

Let's have a look at this data

```{r}
head(raw_data)
dim(raw_data)
```


## What is mean total number of steps taken per day?

First of all, let's take only non-missing values into account and subset a corresponding dataset from the original data

```{r}
subdata <- raw_data[complete.cases(raw_data), ]
```

Let's see inside of this subset

```{r}
head(subdata)
dim(subdata)
```


Now we make a histogram of the total number of steps taken each day

```{r}
totals_days <- aggregate(steps ~ date, subdata, sum)

hist(totals_days$steps,  col = "red", main = "Total Steps Per Day", 
     xlab = "Total Number of Dayly Steps", ylab = "Number of Days")
```

Calculate and report the mean and median total number of steps taken per day

```{r}
mean(totals_days$steps)
median(totals_days$steps)
```

## What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
totals_intervals <- aggregate(steps ~ interval, subdata, mean)
plot(totals_intervals$interval, totals_intervals$steps, type="l",
     xlab="Number of interval", ylab = "Average number of steps")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
totals_intervals$interval[which.max(totals_intervals$steps)]
max(totals_intervals$steps)
```

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
sum(!complete.cases(raw_data))
```

Fill in all of the missing values in the dataset using the mean for 5-minute intervals (calculated above).

```{r}
fill_steps <- numeric()
for (i in 1:nrow(raw_data)) {
       observation <- raw_data[i, ]
       if (is.na(observation$steps)) {
             steps <- totals_intervals[totals_intervals$interval == 
                                         observation$interval,]$steps
         } else {
               steps <- observation$steps
           }
       fill_steps <- c(fill_steps, steps)
   }
```

Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
processed_data <- raw_data
processed_data$steps <- fill_steps
```

Make a histogram of the total number of steps taken each day.

```{r}
totals_days_new <- aggregate(steps ~ date, processed_data, sum)
hist(totals_days_new$steps,  col = "red", main = "Total Steps Per Day", 
     xlab = "Total Number of Dayly Steps", ylab = "Number of Days")
```

Calculate and report the mean and median total number of steps taken per day

```{r}
mean(totals_days_new$steps)
median(totals_days_new$steps)
```

As we can see, the median total number of dayly steps has changed and has become equal to
the mean.

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
Sys.setlocale("LC_TIME", "English")
Sys.setenv(TZ = "UTC")
processed_data$date <- strptime(processed_data$date,
               "%Y-%m-%d", tz = "EST5EDT")
processed_data$day <- ifelse(weekdays(processed_data$date) 
                             %in% c("Saturday", "Sunday"),'weekend','weekday')
```

Look at the new processed_data

```{r}
head(processed_data)
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
totals_intervals_dayweeks <- aggregate(steps ~ interval + day,processed_data, mean)

library(lattice)
xyplot(steps ~ interval | day, totals_intervals_dayweeks, type = "l", layout = c(1, 2), 
       xlab = "Interval", ylab = "Number of steps")
```

As we can see from above plot there is a definite difference between activity patterns for
weekdays and weekends.